---
title: "Causal Inference Notebook"
author: "ABS"
date: "2023-12-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction of simple causal questions 

```{r}
library(broom)
library(tidyverse)
library(magrittr)
# getting the data
data <- readr::read_csv(file = "https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/1268/20/nhefs.csv") %>% 
  mutate(
    education = case_when(
      education == 1  ~ "8th grade or less",
      education == 2 ~ "HS dropout",
      education == 3 ~ "HS",
      education == 4 ~ "College dropout",
      education == 5 ~ "College or more",
      T ~ "missing"
    )
  ) %>% 
  mutate(across(.cols = c(qsmk, sex, race, education, exercise, active), .fns = forcats::as_factor)) %>% 
  drop_na(qsmk, sex, race, education, exercise, active, wt82)
data %>% head()

```

## Calculate  Causal effect manually 

Causal effect formula 

$$
RR = E(Y|A=1) - E(Y|A=0) \\
OR = \frac{E(Y|A=1)}{E(Y|A=0)}

$$

```{r}
meanvalue  = data %>% 
  group_by(qsmk) %>% 
  summarise("meanvalue"= mean(wt82_71)) 
  

Causal_effect = md$meanvalue[2]-md$meanvalue[1]  
Causal_effect

```


## Calculate causal effect from a model

The estimate value of treatment is the causal effect of the model

```{r}

mm=lm(wt82_71 ~ qsmk, data = data) %>% broom::tidy(.,conf.int = T) %>% 
  select (estimate, conf.low,conf.high)

mm
```


## IP weighting 

IP weighting is another way of adjusting the treatment value as sometimes values are unadjusted.
Steps :
 * Fit a model of treatment by adjusting the other confounder variables.
 * Then fit the outcome model by weighting the value.
 
Stabilized IP weight vs IP weight 

*Stabilized ip weight = $\frac{E{[A]}}{E[A|L]}$ for treatment 
*Stabilized ip weight = $\frac{1}{E[A|L]}$ for treatment 
 
```{r}
prop.table(table(data$qsmk))
``` 

The table shows that treatment assignment is unbalenced.So we cannot find the actual causal impact for here.So creat a pseudo population and balancing it.

```{r}

treat_model = glm(qsmk ~ 1, family = binomial(link = "logit"), data = data)
treat_model_1 =  glm(data = data, formula = qsmk ~ sex + race + age + I(age*age) + education + smokeintensity + I(smokeintensity*smokeintensity) + smokeyrs + I(smokeyrs*smokeyrs) + exercise + active + wt71 + I(wt71*wt71), family = binomial("logit"))

data = data %>% 
  mutate(
    treat_model_r = predict(treat_model, type = "response"),
    treat_model_1_r = predict(treat_model_1, type = "response"),
    ipw = ifelse(qsmk==1, 1/treat_model_1_r, 1/(1- treat_model_1_r)),
    sbipw = ifelse(qsmk==1,treat_model_r/ treat_model_1_r,(1-treat_model_r)/(1-treat_model_1_r ))
  )
```


### Graph


```{r}
data %>% 
  group_by(qsmk) %>% 
  ggplot(aes(x = ipw, color = qsmk, fill = qsmk)) +
  geom_histogram(bins = 50) +
  theme_minimal() 


```

```{r}
data %>% 
  group_by(qsmk) %>% 
  ggplot(aes(x = sbipw, color = qsmk, fill = qsmk)) +
  geom_histogram(bins = 50) +
  theme_minimal() 

```



### fit the outcome model for unstabilized method

```{r}
om = glm(wt82_71 ~  qsmk ,weights = ipw, data = data,family = gaussian())
om %>% tidy()

```

### Fit the causal model for stabilized method

```{r}
om = glm(wt82_71 ~  qsmk ,weights = sbipw, data = data,family = gaussian())
om %>% tidy()

```


### Bootstrappning

```{r}

library(tidymodels)
set.seed(972188635)
boots <- bootstraps(data, times = 1e3, apparent = FALSE)
iptw_model_sw <- function(data) {
  glm(formula = wt82_71 ~ qsmk, family = gaussian(), weight = sbipw, data = data)
}
boot_models <- boots %>% 
  mutate(
    model = map(.x = splits, ~iptw_model_sw(data = .x)),
    coef_info = map(model, ~broom::tidy(.x)))
int_pctl(boot_models, coef_info)
```


### Checking positivity

* Structural violations
  - Individuals with some levels of confounders cannot be treated (contraindcations for medications, etc.)
  - Causal inferences are impossible for the entire population using IPW or standardization
* Random violations
  - Finite data sample
  - Parametric modeling to smooth over random zeroes
  
  
### Marginal structural models (MSMs)

* Models for marginal potential outcomes
* Marginal: marginal effect of exposure on the outcome
* Structural: for PO

```{r}
data_smkint <- data %>% filter(smokeintensity < 26)
data_smkint %>% head()
```

## Fitting IP weighting 

* Here is for continuous treatment.

```{r}
pr_a <- lm(data = data_smkint, formula = smkintensity82_71~1)
pr_a_l <- lm(data = data_smkint, formula = smkintensity82_71 ~ sex + race + age + I(age*age) + education + smokeintensity + I(smokeintensity*smokeintensity) + smokeyrs + I(smokeyrs*smokeyrs) + exercise + active + wt71 + I(wt71*wt71))
sd_a <- summary(pr_a)$sigma
sd_a_l <- summary(pr_a_l)$sigma
data_smkint %<>% mutate(
  p_a = predict(object = pr_a, type = "response"),
  p_a_l = predict(object = pr_a_l, type = "response"),
  ## continuous outcome 
  density_num = dnorm(x = smkintensity82_71, mean = p_a, sd = sd_a),
  density_denom = dnorm(x = smkintensity82_71, mean = p_a_l, sd = sd_a_l),
  # for average treatment effect
  sw_iptw = density_num/density_denom
)
summary(data_smkint$sw_iptw)
```

### Bootstrapping 

```{r}
set.seed(198877)
boots <- bootstraps(data_smkint, times = 1e3, apparent = FALSE)
iptw_model <- function(data) {
  glm(formula = wt82_71 ~ smkintensity82_71 + I(smkintensity82_71*smkintensity82_71), family = gaussian(), weight = sw_iptw, data = data)
}
boot_models <- boots %>% 
  mutate(
    model = map(.x = splits, ~iptw_model(data = .x)),
    coef_info = map(model, ~broom::tidy(.x)))
int_pctl(.data = boot_models, statistics = coef_info)
```


### Now stratification based on sex 


```{r}
treat_model_s = glm(qsmk ~ sex, family = binomial(link = "logit"), data = data)
treat_model_1_s =  glm(data = data, formula = qsmk ~ sex + race + age + I(age*age) + education + smokeintensity + I(smokeintensity*smokeintensity) + smokeyrs + I(smokeyrs*smokeyrs) + exercise + active + wt71 + I(wt71*wt71), family = binomial("logit"))

data = data %>% 
  mutate(
    treat_model_r = predict(treat_model_s, type = "response"),
    treat_model_1_r = predict(treat_model_1_s, type = "response"),
    ipw = ifelse(qsmk==1, 1/treat_model_1_r, 1/(1- treat_model_1_r)),
    sbipw = ifelse(qsmk==1,treat_model_r/ treat_model_1_r,(1-treat_model_r)/(1-treat_model_1_r ))
  )
```



```{r}
set.seed(972188635)
boots <- bootstraps(data, times = 1e3, apparent = FALSE)
iptw_model_sw <- function(data) {
  glm(formula = wt82_71 ~ qsmk*sex, family = gaussian(), weight = sbipw, data = data)
}
boot_models <- boots %>% 
  mutate(
    model = map(.x = splits, ~iptw_model_sw(data = .x)),
    coef_info = map(model, ~broom::tidy(.x)))
int_pctl(boot_models, coef_info)

```


## Censoring and missing data

```{r}
data <- readr::read_csv(file = "https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/1268/20/nhefs.csv") %>% 
  mutate(
    education = case_when(
      education == 1  ~ "8th grade or less",
      education == 2 ~ "HS dropout",
      education == 3 ~ "HS",
      education == 4 ~ "College dropout",
      education == 5 ~ "College or more",
      T ~ "missing"
    ),
    cens = case_when(
      is.na(wt82) ~ 1, 
      T ~ 0
    )
  ) %>% 
  mutate(across(.cols = c(qsmk, sex, race, education, exercise, active, cens), .fns = forcats::as_factor))

```

### IP weigthing for normal cases

```{r}
pr_a <- glm(data = data, formula = qsmk~1, family = binomial("logit"))
pr_a_l <- glm(data = data, formula = qsmk ~ sex + race + age + I(age*age) + education + smokeintensity + I(smokeintensity*smokeintensity) + smokeyrs + I(smokeyrs*smokeyrs) + exercise + active + wt71 + I(wt71*wt71), family = binomial("logit"))
data %<>% mutate(
  p_a = predict(object = pr_a, type = "response"),
  p_a_l = predict(object = pr_a_l, type = "response"),
  sw_iptw = if_else(qsmk == 1, p_a/p_a_l, (1-p_a)/(1-p_a_l)) # stabilized weights
)
```

### Ip weighting for censoring

```{r}
pr_c <- glm(data = data, formula = cens~qsmk, family = binomial("logit"))
pr_c_l <- glm(data = data, formula = cens ~ qsmk + sex + race + age + I(age*age) + education + smokeintensity + I(smokeintensity*smokeintensity) + smokeyrs + I(smokeyrs*smokeyrs) + exercise + active + wt71 + I(wt71*wt71), family = binomial("logit"))
data %<>% mutate(
  p_c = predict(object = pr_c, type = "response"),
  p_c_l = predict(object = pr_c_l, type = "response"),
  sw_ipcw = if_else(cens == 0, (1-p_c)/(1-p_c_l), 1),
  w = sw_iptw * sw_ipcw
)

```

### Causal effect for censoring 

```{r}
set.seed(9721835)
boots <- bootstraps(data, times = 1e3, apparent = FALSE)
w_model <- function(data) {
  glm(formula = wt82_71 ~ qsmk, family = gaussian(), weight = w, data = data)
}
boot_models <- boots %>% 
  mutate(
    model = map(.x = splits, ~w_model(data = .x)),
    coef_info = map(model, ~broom::tidy(.x)))
int_pctl(boot_models, coef_info)

```



# Chapter 2 Standardization 


